---
title: "Week 3 In-class Assignment"
author: "Khawlah Alshammari"
format: 
  pdf:
    geometry:
      - top=30mm
      - left=20mm
      - heightrounded
---

```{r}
#| output: false
#| warning: false
#| message: false
library(tidyverse)
library(knitr)
library(kableExtra)
library(dplyr)
library(janitor)
```

**Prepare World Bank data**

1- Reading the data. First we read in maternal mortality data, then changing to long format.

```{r}
# Read in maternal mortality data
matmor <- read.csv("../data/raw/maternal_mortality.csv", header = TRUE)

# Change wide to long format
matmor_long <- matmor |>
  pivot_longer(cols = starts_with("X"),
                 names_to = "year",
                 names_prefix = "X", # removes X from year column
                 values_to = "maternal_mortality") |> 
  mutate(year = as.numeric(year)) |> # change year to numeric
  select(iso, year, maternal_mortality)

```


2- Creating a function that can be used to convert all data sets we have into long formats.


```{r}
mortality_fun <- function(data, varname) {

    data |>
        pivot_longer(
            cols = starts_with("X"),
            names_to = "year",
            names_prefix = "X",
            values_to = varname
        ) |>
        mutate(year = as.numeric(year)) |>
        select(iso, year, all_of(varname))
}
```


```{r}
infmor <- read.csv("../data/raw/infant_mortality.csv", header = TRUE)

infmor_long <- mortality_fun(
    data = infmor,
    varname = "infant_mortality"
)

neomor <- read.csv(("../data/raw/neonatal_mortality.csv"), header = TRUE)

neomor_long <- mortality_fun(
    data = neomor,
    varname = "neonatal_mortality"
)

un5mor <- read.csv(("../data/raw/under5_mortality.csv"), header = TRUE)

un5mor_long <- mortality_fun(
    data = un5mor,
    varname = "under5_mortality"
)

```


**Prepare disaster data**

1- Reading the disaster data.

```{r}
disaster <- read.csv("../data/raw/disaster.csv", header = TRUE)
```

2- Cleaning the variable names, then including only data for 2000-2019, and finally creating th dummey variable that counts droughts and earthquakes.


```{r}
disaster_clean <- disaster |>
    clean_names() |>
    filter(
        year >= 2000,
        year <= 2019,
        disaster_type %in% c("Earthquake", "Drought")
    ) |>
    select(year, iso, disaster_type) |>
    mutate(
        earthquake = if_else(disaster_type == "Earthquake",1,0),
        drought = if_else(disaster_type == "Drought", 1, 0)
    ) |>
    group_by(year,iso) |>
    summarise(
        earthquake = max(earthquake),
        drought = max(drought),
        .groups = "drop"
    )
```


**Prepare conflict data**

1- Reading the conflict data set.

```{r}
conflict <- read.csv("../data/raw/conflict.csv", header = TRUE)
```

2- Creating the binary armed conflict vatiable.

```{r}
conflict_clean <- conflict |>
  mutate(
    armed_conflict = if_else(best >= 25, 1, 0)
    # binary variable indicating the presence of conflict for each country–year observation (0 = no, <25 battle-related deaths; 1 = yes, >=25 battle-related deaths)
  ) |>
  group_by(iso, year) |>
  summarise(
    armed_conflict = as.integer(max(armed_conflict)),
    .groups = "drop"
  )

conflict_clean <- conflict_clean |>
  mutate(
    year = as.integer(year+1)
  ) |>
  ungroup()
```

**Merging all data sets**

```{r}
final_data <- list(
  matmor_long,
  infmor_long,
  neomor_long,
  un5mor_long,
  disaster_clean,
  conflict_clean
) |>
  reduce(full_join, by = c("iso", "year"))

write.csv(final_data,"../data/processed/final_data.csv", row.names=FALSE)
```